---
title: "20231027 MS analysis of Co-IP"
output: html_document
date: "2023-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo=FALSE, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
   #  install.packages("BiocManager")
# BiocManager::install()

library(DEP)
library(dplyr)
library(shiny)
library(tidyr)
```
# load data and remove contaminants
```{r data_contaminants}
#load data
my_data <- read.delim("proteinGroups.txt")
data <-my_data
experimental_design <- read.delim("20231027_experiment_all.txt")

# We filter for contaminant proteins and decoy database hits, which are indicated by "+" in the columns "Potential.contaminants" and "Reverse", respectively. 
data <- filter(data, Reverse != "+", Potential.contaminant != "+")


#Are there any duplicated gene names?
data$Gene.names %>% duplicated() %>% any()

#Make a table of duplicated gene names
data %>% group_by(Gene.names) %>% summarize(frequency = n()) %>% 
  arrange(desc(frequency)) %>% filter(frequency > 1)

#For further analysis these proteins must get unique names. Additionally,
#some proteins do not have an annotated gene name and for those we will use the Uniprot ID.

#Make unique names using the annotation in the "Gene.names" column as primary names and the annotation in "Protein.IDs" as name for those that do not have an gene name.
data_unique <- make_unique(data, "Gene.names", "Protein.IDs", delim = ";")

#Are there any duplicated names?
data$name %>% duplicated() %>% any()

#Generate a SummarizedExperiment object using an experimental design
LFQ_columns <- grep("LFQ.", colnames(data_unique)) # get LFQ column numbers
data_se <- make_se(data_unique, LFQ_columns, experimental_design)

#Generate a SummarizedExperiment object by parsing condition information from the column names
LFQ_columns <- grep("LFQ.", colnames(data_unique)) # get LFQ column numbers
data_se_parsed <- make_se_parse(data_unique, LFQ_columns)

#Let's have a look at the SummarizedExperiment object
data_se
```
# Filter for occurence in replicates
```{r filt}
# Plot a barplot of the protein identification overlap between samples
plot_frequency(data_se)

#filter_missval filters a proteomics dataset based on missing values.
#The dataset is filtered for proteins that have a maximum of ’thr’ missing values in at least one condition.

#IF filter for proteins that are identified in all replicates of at least one condition:
#data_filt <- filter_missval(data_se, thr = 0)

# OR FOR Less stringent filtering:
#Filter for proteins that are identified in 3 out of 6 replicates of at least one condition
data_filt <- filter_missval(data_se, thr = 3)


#After filtering, the number of identified proteins per sample can be plotted as well as the overlap in identifications between samples.

# Plot a barplot of the number of identified proteins per samples
plot_numbers(data_filt)

# Plot a barplot of the protein identification overlap between samples
plot_coverage(data_filt)

```

# Test normalization of data and decide with plot whether to use it
```{r norm}
#The data is background corrected and normalized by variance stabilizing transformation (vsn).

# Normalize the data
data_norm <- normalize_vsn(data_filt)
#The normalization can be inspected by checking the distributions of the samples before and after normalization.

# Visualize normalization by boxplots for all samples before and after normalization
plot_normalization(data_filt, data_norm)
#-> looks worse with normalization than without I think, so don't normalize, however GFP linker replicates seem badly aligned in data_fil
```

# WITHOUT normalization
## Imputation 
```{r imp}
#Plot a heatmap of proteins with missing values
plot_missval(data_filt)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
plot_detect(data_filt)


data_imp <- impute(data_filt, fun = "min") #without norm, if want norm, need to impute "data_norm"

# if want a different method, all possible imputation methods are printed in an error, if an invalid function name is given.
#impute(data_filt, fun = "")
#more explanation: https://rdrr.io/bioc/MSnbase/man/impute-methods.html

plot_imputation(data_filt, data_imp) #without norm, if want norm, need to impute "data_norm"

```
## Differential enrichment analysis  based on linear models and empherical Bayes statistics
### parental vs rest LFC = 1.5
```{r fig.width=4, fig.height=4}
# Test every sample versus control
data_diff <- test_diff(data_imp, type = "control", control = "parental")
##Tested contrasts: GFP_link_vs_parental, cand1_vs_parental, cand2_vs_parental, cand11_vs_parental, cand12_vs_parental, cand16_vs_parental, cand19_vs_parental
 
#otherwise also possible to test all possible comparisons of samples:
#data_diff_all_contrasts <- test_diff(data_imp, type = "all")

#or to test manually defined comparisons
#data_diff_manual <- test_diff(data_imp, type = "manual", 
                         #     test = c("RPL11_vs_cand1", "cand1_vs_RPL11"))


##Finally, significant proteins are defined by user-defined cutoffs using add_rejections.

# Denote significant proteins based on user defined cutoffs, vs control
dep <- add_rejections(data_diff, alpha = 0.05, lfc = log2(1.5))
#OR
# Denote significant proteins based on user defined cutoffs, vs all
#dep <- add_rejections(data_diff_all_contrasts, alpha = 0.05, lfc = log2(1.5)) 

# Plot the first and second principal components
plot_pca(dep, x = 1, y = 2, n = 54, point_size = 4)

# Plot the Pearson correlation matrix
plot_cor(dep, significant = TRUE, lower = 0, upper = 1, pal = "Reds")

# Plot a heatmap of all significant proteins with the data centered per protein
plot_heatmap(dep, type = "centered", kmeans = TRUE, 
             k = 3, col_limit = 4, show_row_names = FALSE,
             indicate = c("condition", "replicate"))

# Plot a heatmap of all significant proteins (rows) and the tested contrasts (columns)
plot_heatmap(dep, type = "contrast", kmeans = TRUE, 
             k = 3, col_limit = 10, show_row_names = FALSE)

# Plot a volcano plot for the contrast " vs Ctrl""
plot_volcano(dep, contrast = "GFP_link_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep, contrast = "cand1_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep, contrast = "cand2_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep, contrast = "cand11_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep, contrast = "cand12_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep, contrast = "cand16_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep, contrast = "cand19_vs_parental", label_size = 2, add_names = TRUE)

pdf('20231027_allvsparental_filt3out6reps_NOnorm_imp_min_lfc_1_5.pdf',width = 4, height = 4) 
plot_volcano(dep, contrast = "GFP_link_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("Fig 3F")
plot_volcano(dep, contrast = "cand1_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("Additional Replicate")
plot_volcano(dep, contrast = "cand2_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("Fig 3B")
plot_volcano(dep, contrast = "cand11_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("Fig 3C")
plot_volcano(dep, contrast = "cand12_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("Fig 3D")
plot_volcano(dep, contrast = "cand16_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("SFig 4C")
plot_volcano(dep, contrast = "cand19_vs_parental", label_size = 2, add_names = TRUE) + ggtitle("Fig 3E")
dev.off()

# Generate a results table
data_results <- get_results(dep)

#save the results table
write.table(data_results, "20231027_allvsparental_filt3out6reps_NOnorm_imp_min_lfc_1_5.txt", sep="\t")
```

```{r}
write.table(data_results[data_results$cand1_vs_parental_significant & data_results$cand1_vs_parental_ratio>1.5,c(1,2,grep("cand1_",colnames(data_results)))],"20231027_cand1.txt", sep="\t", quote = F) 

write.table(data_results[data_results$cand2_vs_parental_significant & data_results$cand2_vs_parental_ratio>1.5,c(1,2,grep("cand2_",colnames(data_results)))],"20231027_cand2.txt", sep="\t", quote = F)

write.table(data_results[data_results$cand11_vs_parental_significant & data_results$cand11_vs_parental_ratio>1.5,c(1,2,grep("cand11_",colnames(data_results)))],"20231027_cand11.txt", sep="\t", quote = F)

write.table(data_results[data_results$cand12_vs_parental_significant & data_results$cand12_vs_parental_ratio>1.5,c(1,2,grep("cand12_",colnames(data_results)))],"20231027_cand12.txt", sep="\t", quote = F)

write.table(data_results[data_results$cand16_vs_parental_significant & data_results$cand16_vs_parental_ratio>1.5,c(1,2,grep("cand16_",colnames(data_results)))],"20231027_cand16.txt", sep="\t", quote = F)

write.table(data_results[data_results$cand19_vs_parental_significant & data_results$cand19_vs_parental_ratio>1.5,c(1,2,grep("cand19_",colnames(data_results)))],"20231027_cand19.txt", sep="\t", quote = F)
```

```{r manual collector}
all.sig <- list()
all.sig$cand1.2 <- data_results[data_results$cand1_vs_parental_significant & data_results$cand1_vs_parental_ratio>1.5,1]
all.sig$cand2.2 <- data_results[data_results$cand2_vs_parental_significant & data_results$cand2_vs_parental_ratio>1.5,1]
all.sig$cand11.2 <- data_results[data_results$cand11_vs_parental_significant & data_results$cand11_vs_parental_ratio>1.5,1]
all.sig$cand12.2 <- data_results[data_results$cand12_vs_parental_significant & data_results$cand12_vs_parental_ratio>1.5,1]
all.sig$cand16.2 <- data_results[data_results$cand16_vs_parental_significant & data_results$cand16_vs_parental_ratio>1.5,1]
all.sig$cand19.2 <- data_results[data_results$cand19_vs_parental_significant & data_results$cand19_vs_parental_ratio>1.5,1]
all.sig$link.2 <- data_results[data_results$GFP_link_vs_parental_significant & data_results$GFP_link_vs_parental_ratio>1.5,1]
saveRDS(all.sig, file="list_all.sig.RData")
```

### parental vs rest LFC = 1
```{r test_2}
# Denote significant proteins based on user defined cutoffs, vs control
dep_2 <- add_rejections(data_diff, alpha = 0.05, lfc = log2(1))
#OR
# Denote significant proteins based on user defined cutoffs, vs all
#dep <- add_rejections(data_diff_all_contrasts, alpha = 0.05, lfc = log2(1.5)) 

# Plot the first and second principal components
plot_pca(dep_2, x = 1, y = 2, n = 54, point_size = 4)

# Plot the Pearson correlation matrix
plot_cor(dep_2, significant = TRUE, lower = 0, upper = 1, pal = "Reds")

# Plot a heatmap of all significant proteins with the data centered per protein
plot_heatmap(dep_2, type = "centered", kmeans = TRUE, 
             k = 3, col_limit = 4, show_row_names = FALSE,
             indicate = c("condition", "replicate"))

# Plot a heatmap of all significant proteins (rows) and the tested contrasts (columns)
plot_heatmap(dep_2, type = "contrast", kmeans = TRUE, 
             k = 3, col_limit = 10, show_row_names = FALSE)

# Plot a volcano plot for the contrast " vs Ctrl""
#pdf('20231027_allvsparental_filt3out6reps_NOnorm_imp_min_lfc_1.pdf')
plot_volcano(dep_2, contrast = "GFP_link_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_2, contrast = "cand1_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_2, contrast = "cand2_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_2, contrast = "cand11_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_2, contrast = "cand12_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_2, contrast = "cand16_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_2, contrast = "cand19_vs_parental", label_size = 2, add_names = TRUE)
#dev.off()

# Generate a results table
data_results_2 <- get_results(dep_2)

#save the results table
#write.table(data_results_2, "20231027_allvsparental_filt3out6reps_NOnorm_imp_min_lfc_1.txt", sep="\t")
```


# WITH normalization
## Imputation 
```{r imp norm}
#Plot a heatmap of proteins with missing values
plot_missval(data_norm)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
plot_detect(data_norm)


data_imp_norm <- impute(data_norm, fun = "min") 

# if want a different method, all possible imputation methods are printed in an error, if an invalid function name is given.
#impute(data_filt, fun = "")
#more explanation: https://rdrr.io/bioc/MSnbase/man/impute-methods.html

plot_imputation(data_norm, data_imp_norm) 

```
## Differential enrichment analysis  based on linear models and empherical Bayes statistics
### parental vs rest LFC = 1.5
```{r test_norm}
# Test every sample versus control
data_diff_norm <- test_diff(data_imp_norm, type = "control", control = "parental")
##Tested contrasts: GFP_link_vs_parental, cand1_vs_parental, cand2_vs_parental, cand11_vs_parental, cand12_vs_parental, cand16_vs_parental, cand19_vs_parental
 
#otherwise also possible to test all possible comparisons of samples:
#data_diff_all_contrasts <- test_diff(data_imp, type = "all")

#or to test manually defined comparisons
#data_diff_manual <- test_diff(data_imp, type = "manual", 
                         #     test = c("RPL11_vs_cand1", "cand1_vs_RPL11"))


##Finally, significant proteins are defined by user-defined cutoffs using add_rejections.

# Denote significant proteins based on user defined cutoffs, vs control
dep_norm <- add_rejections(data_diff_norm, alpha = 0.05, lfc = log2(1.5))
#OR
# Denote significant proteins based on user defined cutoffs, vs all
#dep <- add_rejections(data_diff_all_contrasts, alpha = 0.05, lfc = log2(1.5)) 

# Plot the first and second principal components
plot_pca(dep_norm, x = 1, y = 2, n = 54, point_size = 4)

# Plot the Pearson correlation matrix
plot_cor(dep_norm, significant = TRUE, lower = 0, upper = 1, pal = "Reds")

# Plot a heatmap of all significant proteins with the data centered per protein
plot_heatmap(dep_norm, type = "centered", kmeans = TRUE, 
             k = 3, col_limit = 4, show_row_names = FALSE,
             indicate = c("condition", "replicate"))

# Plot a heatmap of all significant proteins (rows) and the tested contrasts (columns)
plot_heatmap(dep_norm, type = "contrast", kmeans = TRUE, 
             k = 3, col_limit = 10, show_row_names = FALSE)

# Plot a volcano plot for the contrast " vs Ctrl""
plot_volcano(dep_norm, contrast = "GFP_link_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm, contrast = "cand1_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm, contrast = "cand2_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm, contrast = "cand11_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm, contrast = "cand12_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm, contrast = "cand16_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm, contrast = "cand19_vs_parental", label_size = 2, add_names = TRUE)

#pdf('20231027_allvsparental_filt3out6reps_norm_imp_min_lfc_1_5.pdf')
#plot_volcano(dep_norm, contrast = "GFP_link_vs_parental", label_size = 2, add_names = TRUE)
#plot_volcano(dep_norm, contrast = "cand1_vs_parental", label_size = 2, add_names = TRUE)
#plot_volcano(dep_norm, contrast = "cand2_vs_parental", label_size = 2, add_names = TRUE)
#plot_volcano(dep_norm, contrast = "cand11_vs_parental", label_size = 2, add_names = TRUE)
#plot_volcano(dep_norm, contrast = "cand12_vs_parental", label_size = 2, add_names = TRUE)
#plot_volcano(dep_norm, contrast = "cand16_vs_parental", label_size = 2, add_names = TRUE)
#plot_volcano(dep_norm, contrast = "cand19_vs_parental", label_size = 2, add_names = TRUE)
#dev.off()

# Generate a results table
data_results_norm <- get_results(dep_norm)

#save the results table
#write.table(data_results_norm, "20231027_allvsparental_filt3out6reps_norm_imp_min_lfc_1_5.txt", sep="\t")
```

### parental vs rest LFC = 1
```{r test_2_norm}
# Denote significant proteins based on user defined cutoffs, vs control
dep_norm_2 <- add_rejections(data_diff_norm, alpha = 0.05, lfc = log2(1))
#OR
# Denote significant proteins based on user defined cutoffs, vs all
#dep <- add_rejections(data_diff_all_contrasts, alpha = 0.05, lfc = log2(1.5)) 

# Plot the first and second principal components
plot_pca(dep_norm_2, x = 1, y = 2, n = 54, point_size = 4)

# Plot the Pearson correlation matrix
plot_cor(dep_norm_2, significant = TRUE, lower = 0, upper = 1, pal = "Reds")

# Plot a heatmap of all significant proteins with the data centered per protein
plot_heatmap(dep_norm_2, type = "centered", kmeans = TRUE, 
             k = 3, col_limit = 4, show_row_names = FALSE,
             indicate = c("condition", "replicate"))

# Plot a heatmap of all significant proteins (rows) and the tested contrasts (columns)
plot_heatmap(dep_norm_2, type = "contrast", kmeans = TRUE, 
             k = 3, col_limit = 10, show_row_names = FALSE)

# Plot a volcano plot for the contrast " vs Ctrl""
#pdf('20231027_allvsparental_filt3out6reps_norm_imp_min_lfc_1.pdf')
plot_volcano(dep_norm_2, contrast = "GFP_link_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm_2, contrast = "cand1_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm_2, contrast = "cand2_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm_2, contrast = "cand11_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm_2, contrast = "cand12_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm_2, contrast = "cand16_vs_parental", label_size = 2, add_names = TRUE)
plot_volcano(dep_norm_2, contrast = "cand19_vs_parental", label_size = 2, add_names = TRUE)
#dev.off()

# Generate a results table
data_results_norm_2 <- get_results(dep_norm_2)

#save the results table
#write.table(data_results_norm_2 , "20231027_allvsparental_filt3out6reps_norm_imp_min_lfc_1.txt", sep="\t")
```